```{r loadEdSurvey6, echo=FALSE, message=FALSE}
library(EdSurvey)
sdf <- readNAEP(path = system.file("extdata/data", "M36NT2PM.dat", package = "NAEPprimer"))
```

# Data Manipulation in `EdSurvey` and Base R {#dataManipulation}

`EdSurvey` gives users functions to efficiently analyze education survey data. As mentioned in [Chapter 3](#philosophyOfAnalysis), `EdSurvey` is flexible when conducting data manipulation and preparing for survey analysis. It allows for rudimentary data manipulation and analysis with both `EdSurvey` and base R functions to edit data before processing. Also, by calling the function `getData()`, one can extract a `light.edsurvey.data.frame` to manipulate similarly as other `data.frame` objects in other R packages. This concept is further detailed in [Chapter 9](#analysisOutsideEdSurvey)analysisOutsideEdSurvey).

## Subsetting the Data

Analysts can directly use a subset of a dataset with `EdSurvey` functions. In this example, a summary table is created with `edsurveyTable` after filtering the sample to include only those students whose value for the `dsex` variable is male and race (as the variable `sdracem`) is either 1 or 3 (White or Hispanic). Both value levels and labels can be used in `EdSurvey` functions.

```{r subset6, cache=FALSE, warning=FALSE}
sdf <- readNAEP(path = system.file("extdata/data", "M36NT2PM.dat", package = "NAEPprimer"))
sdfm <- subset(x = sdf, subset = dsex == "Male" & (sdracem == 3 | sdracem == 1))
es1 <- edsurveyTable(formula = composite ~ dsex + sdracem, data = sdfm)
```
```{r es1, eval=FALSE}
es1
```
```{r table601, echo=FALSE}
knitr::kable(x = es1$data, digits = 7, row.names = FALSE, caption = "Subsetting Data \\label{tab:subsettingData}")
```


## Recoding Variable Names and Levels Using `recode.sdf` and `rename.sdf`

To assist in the process of standardizing data for an `edsurvey.data.frame`, a `light.edsurvey.data.frame`, and an `edsurvey.data.frame.list`, the functions `recode.sdf()` and `rename.sdf()` are very handy. 

The `recode.sdf()` function accepts the levels of a variable as a vector from their current values to their newly recoded value. For example, changing the lowest level of `b017451` from `"Never or hardly ever"` to `"Infrequently"` and the highest level from `"Every day"` to `"Frequently"`, will recode levels for that variable in our connection to `sdf`:

```{r recode}
sdf2 <- recode.sdf(sdf, recode = list(
  b017451 = list(from = c("Never or hardly ever"),
               to = c("Infrequently")),
  b017451 = list(from = c("Every day"),
               to = c ("Frequently"))
  )
)
searchSDF(string = "b017451", data = sdf2, levels = TRUE)                              
```

In addition, we can change the name of variables using `rename.sdf()`. The recoded variable `"b017451"` can be changed to a value that more effectively describes its contents, such as `"studytalkfrequency"`:

```{r rename}
sdf2 <- rename.sdf(x = sdf2, oldnames = "b017451", newnames = "studytalkfrequency")
searchSDF(string = "studytalkfrequency", data = sdf2, levels = TRUE)
```

Most `EdSurvey` analytical functions allow the user to recode variable levels through the recode argument, including, for example, `cor.sdf()`, `lm.sdf()`, and `edsurveyTable()`. Reference a function's documentation page for details.

It also is important to note that `EdSurvey` functions (and function arguments) do not permanently overwrite the variable information from your data source; they recode it only for the current connection to the data in R. The original file formatting can be retrieved by reconnecting to the data source via `readNAEP()`.


## Retrieving Data for Further Manipulation With `getData`

Data can be extracted and manipulated using the function `getData`. The function `getData` takes an `edsurvey.data.frame` and returns a `light.edsurvey.data.frame` containing the requested variables by either specifying a set of variable names in `varnames` or entering a formula in `formula`.[^helpgetData] For plausible values and weights, only the names of the main scale/subscale and the weight variable need to be included here for all necessary supplementary variables to be automatically included.

To access and manipulate data for the `dsex` and `b017451` variables and the `num_oper` subject scale in `sdf`, call `getData`. In the following code, the `head` function is used, which reveals only the first few rows of the resulting data:

```{r getData, warning=FALSE}
gddat <- getData(data = sdf, varnames = c("dsex","b017451", "num_oper"),
                 dropOmittedLevels = TRUE)
head(gddat)
```

By default, setting `dropOmittedLevels` to `TRUE` removes special values such as multiple entries or instances of `NA`. `getData` tries to help by dropping the levels of factors for regression, tables, and correlations that are not typically included in analysis.

## Retrieving All Variables in a Dataset {#retrievingAllVariablesInADataset}

To extract all data in an `edsurvey.data.frame`, define the `varnames` argument as `colnames(sdf)`, which will query all variables. Setting the arguments `dropOmittedLevels` and `defaultConditions` to `FALSE` ensures that values that would normally be removed are included:

```{r lsdf0, eval = FALSE, warning=FALSE}
lsdf0 <- getData(data = sdf, varnames = colnames(sdf), addAttributes = TRUE,
                 dropOmittedLevels = FALSE, defaultConditions = FALSE)
dim(x = lsdf0) # excludes the one school variable in the sdf
dim(x = sdf)
```

Once retrieved, this dataset can be used with all `EdSurvey` functions.

## Using `EdSurvey` Functions on a Unique `light.edsurvey.data.frame`

After manipulating the data, you can use a `light.edsurvey.data.frame` with any `EdSurvey` function. Most notably, a `light.edsurvey.data.frame` can create tables using `edsurveyTable` and run regressions with the `lm.sdf` function.

### `edsurveyTable`
The following example creates an `edsurveyTable` using the manipulated `light.edsurvey.data.frame` (named `gddat`), the variables `dsex` and `b017451`, the five plausible values for `composite`, and the default weight `origwt`:[^qedsurveyTable]

[^qedsurveyTable]: Consult `?edsurveyTable` or the vignette titled [*Using EdSurvey to Analyze NCES Data: An Illustration of Analyzing NAEP Primer*](https://nces.ed.gov/nationsreportcard/subject/researchcenter/pdf/using_edsurvey_for_naep.pdf) for details on default `edsurveyTable` arguments.


```{r edsurveyTableGddat}
gddat <- getData(data = sdf, varnames = c("composite", "dsex", "b017451",
                                          "c052601","origwt"), addAttributes = TRUE)
es2 <- edsurveyTable(formula = composite ~ dsex + b017451,
                     weightVar = "origwt", data = gddat)
```

```{r table602, echo=FALSE}
knitr::kable(x = es2$data, digits = 7, row.names = FALSE, caption = "Using EdSurvey Functions on a light.edsurvey.data.frame \\label{tab:lsdf}")
```

### `lm.sdf`
To generate a linear model using a `light.edsurvey.data.frame`, the included arguments from the previous example, as well as the weight `origwt`, are passed through the `lm.sdf` function:[^qlmsdf]

[^qlmsdf]: Consult `?lm.sdf` or the appendix of the vignette titled *Using EdSurvey to Analyze NCES Data: An Illustration of Analyzing NAEP Primer* for details on default `lm.sdf` arguments.


```{r lmsdfGddat}
lm2 <- lm.sdf(formula = composite ~ dsex + b017451, weightVar = "origwt", data = gddat)
summary(object = lm2)
```

Contrasts from treatment groups also can be omitted from a linear model by stating the variable name in the `relevels` argument. In this example, values with `dsex = "Female"` are withheld from the regression. Use the base R function `summary` to view details about the linear model.

```{r lmsdfGddat2}
lm3 <- lm.sdf(formula = composite ~ dsex + b017451, data = gddat,
              relevels = list(dsex = "Female"))
summary(object = lm3)
```

### cor.sdf
Users might want to generate a correlation to explore a manipulated `light.edsurvey.data.frame`. The marginal correlation coefficient among plausible values of the subject scales and subscales can be calculated on a `light.edsurvey.data.frame` object `eddat` using the `cor.sdf` function and the Pearson method. In this example, the variable `dsex == "Female"` subsets our `light.edsurvey.data.frame` to calculate the correlation between the subject subscales `num_oper` and `algebra` using the default weight `origwt`:[^corsdf]

[^corsdf]: Consult `?cor.sdf` or the appendix of the vignette titled *Using EdSurvey to Analyze NCES Data: An Illustration of Analyzing NAEP Primer* for details on default `cor.sdf` arguments.

```{r corsdfGddat}
eddat <- getData(data = sdf, varnames = c("num_oper","algebra","dsex", 'origwt'),
                addAttributes = TRUE, dropOmittedLevels = FALSE)
eddat <- subset(eddat,dsex == "Female")
cor2 <- cor.sdf(x = "num_oper", y = "algebra", weightVar = "origwt",
                data = eddat, method = "Pearson")
cor2
```

## Applying `rebindAttributes` to Use `EdSurvey` Functions With Manipulated Data Frames

A helper function that pairs well with `getData` is `rebindAttributes`. This function allows users to reassign the attributes from a survey dataset to a data frame that might have had its attributes stripped during the manipulation process. Once attributes have been rebinded, all variables---including those outside the original dataset---are available for use in `EdSurvey` analytical functions.

For instance, the `sdf` object contains the following attributes:

```{r attributes}
attributes(sdf)
```

These attributes are lost when variables are retrieved via `getData()`. For example, a user might want to run a linear model using `composite`, the default weight `origwt`, the variable `dsex`, and the categorical variable `b017451` recoded into a binary variable. To do so, we can return a portion of the `sdf` survey data as the `gddat` object. Next, use the base R function `ifelse` to conditionally recode the variable `b017451` by collapsing the levels `"Never or hardly ever"` and `"Once every few weeks"` into one level (`"Rarely"`) and all other levels into `"At least once a week"`.

```{r rebindAttributesPrep, cache=FALSE, warning=FALSE}
gddat <- getData(data = sdf, varnames = c("dsex", "b017451", "origwt", "composite"),
                 dropOmittedLevels = TRUE)
gddat$studyTalk <- ifelse(gddat$b017451 %in% c("Never or hardly ever",
                                               "Once every few weeks"),
                          "Rarely", "At least once a week")
```

From there, apply `rebindAttributes` from the attribute data `sdf` to the manipulated data frame `gddat`. The new variables are now available for use in `EdSurvey` analytical functions:

```{r rebindAttributeslm, cache=FALSE, warning=FALSE}
gddat <- rebindAttributes(data = gddat, attributeData = sdf)
lm2 <- lm.sdf(formula = composite ~ dsex + studyTalk, data = gddat)
summary(object = lm2)
```

Additional details on the features of the `getData` function appear in the vignette titled [*Using the `getData` Function in EdSurvey*](https://www.air.org/sites/default/files/EdSurvey-getData.pdf).

## Important Data Manipulation Notes

### Memory Usage

Because many NCES databases have hundreds of columns and millions of rows, `EdSurvey` allows users to analyze data without storing it in the global environment. Alternatively, the `getData` function retrieves a `light.edsurvey.data.frame` into the global environment, which can be costlier to memory usage. The base R function `object.size` estimates the memory being used to store an R object. Computations using objects stored in the global environment are markedly costlier to memory than those made directly from the `EdSurvey` database:

```{r memusage}
object.size(gddat <- getData(data = sdf,
                             varnames = c('composite', 'dsex', 'b017451', 'origwt'),
                             addAttributes = TRUE, dropOmittedLevels = FALSE))
object.size(lm7 <- lm.sdf(formula = composite ~ dsex + b017451,
                          weightVar='origwt', data = gddat))
object.size(lm8 <- lm.sdf(formula = composite ~ dsex + b017451,
                          weightVar='origwt', data = sdf))
```

Although a manipulated `light.edsurvey.data.frame` requires nearly 10 MB of working memory to store both the `light.edsurvey.data.frame` and the regression model object (`lm7`), the resulting object of the same computation made directly from the `EdSurvey` database (`lm8`) holds only 5--7 kB. It is a good practice to remove unnecessary values saved in the global environment. Because we have stored many large data objects, let's remove these before moving on.

```{r rm}
rm(df,gddat,eddat)
```

Some operating systems continue to hold the memory usage even after removing an object. R will clean up your global environment automatically, but a forced garbage cleanup also can be employed:

```{r gc, eval = FALSE}
gc()
```

### Forgetting to Include a Column Variable

When creating a summary table or running regression, `EdSurvey` will give a warning when a column is missing:

```{r missingColumns, eval = FALSE}
gddat <- getData(data = sdf, 
                 varnames = c(all.vars(composite ~ lep + dsex + iep), "origwt"), 
                 addAttributes = TRUE, dropOmittedLevels = FALSE)
lm9 <- lm.sdf(formula = composite ~ lep + dsex + iep + b017451, data = gddat)
## Using default weight variable 'origwt'
## Error in getData(sdf, c(all.vars(formula), wgt), ..., includeNaLabel = TRUE)
  ## The following variable names are required for this call 
  ## and are not on the incoming data 'b017451'.
```

The solution is simple: Edit the call to `getData` to include the variable and rerun the linear model.

```{r lmForgotColumn}
gddat <- getData(data = sdf,
                 varnames = c(all.vars(composite ~ lep + dsex + iep + b017451),"origwt"), 
                 addAttributes = TRUE, dropOmittedLevels = FALSE)
lm9 <- lm.sdf(formula = composite ~ lep + dsex + iep + b017451, data = gddat)
lm9
```
